Cartogram Mapping and Application to Cancer Data.

### 2019-05-08 review_article.Rmd
Review article sources contains notes on references and how they are relevant is each section.

@@

Observed, or predicted.
Challenges for either.

>>>>

@MACM
geographically mapping cancer mortality uses maps of europe:
Static choropleth maps, coloured from green to red, supplemented by histograms.
Introduces Moran scatterplot "In the Moran’s scatterplot, each geographic unit is represented by a point, whose X coordinate is the value of the mortality in that area and the Y coordinate is equal to the average value of the theme in the bordering areas'
And Moran's index: spatial correlation with neighbours

'visual intuition coming from the inspection of maps must be complemented and validated by rigorous statistical analyses.'
Does not acknowledge the map as a visual statistic, but a tool.

cites: Cancer Mortality Atlas of EU and of the European Economic Area last published 8 years ago

@@

@SEMA
'Disease mapping is carried out to summarise spatial and spatio-temporal variation in risk.'
Decriptive statistic of the relationship between geography and a disease.
population, exposure, and health data may have pointdata or aggregated summaries (countdata)
potentially be subject to a number of inaccuracies
diagnostic error or misclassification

Not only the counts, but population, migration, deaths and births are modelled across census years.

rapid visual summary of complex geographical information, descriptive purposes, to generate hypotheses, for surveillance, and to aid policy formation and resource allocation.

When using counts:
 small populations imply large variability in estimated rates

Decisions taken at this stage may be critical as apparent ‘clusters’ may depend crucially on the boundaries chosen in time or space: ‘The more narrowly the underlying population is defined, the less will be the number of expected cases, the greater will be the estimate of the excess rate, and often the more pronounced will be the statistical significance’

Using prediction:
assuming that neighbouring μi are correlated leads to a ‘clustering’ prior

### 2019-05-08 review_article_sources.Rmd
Review article sources contains notes on references and how they are relevant is each section.

Utilising maps to present statistics has been widely used, for centuries.
Connecting the data to the geographical representation of areas that are already familiar.
It is not enough to be recognisable, if the value of the statistic cannot be seen. Cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer outcomes. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes.

## Spatial Visualisation of Disease

@@

@@

### counts versus predictions

@SEMA
'Disease mapping is carried out to summarise spatial and spatio-temporal variation in risk.'
Decriptive statistic of the relationship between geography and a disease.
population, exposure, and health data may have pointdata or aggregated summaries (countdata)
potentially be subject to a number of inaccuracies
diagnostic error or misclassification

Not only the counts, but population, migration, deaths and births are modelled across census years.

rapid visual summary of complex geographical information, descriptive purposes, to generate hypotheses, for surveillance, and to aid policy formation and resource allocation.

@@

Driven by progressions in computing, statistical methods and collection of data.

Models summarise relationships between counts of diseases and other factors.

Choropleth Maps, relationship between regions and spatial location. Able to represent all the occurences within areas more cohesively, as total counts, or colours.

Urbanisation pushed people closer together, in workplaces and in homes.

Recognition of biases when utilising the geography to map people. Small areas contain more people, work towards representing regions of importance more clearly.

@@

General introduction:
Cartograms stepped toward this goal by highlighting areas of interest. Made some areas bigger on the map space, kept relationship with neighbours.
USA regular cartogram (UK too)

### Contiguous
Importance of the statistic in this area being communicated was more important than keeping geography recognisable.

@@

## Comparison of 
(only worth including if it is possible for us to implement)
Tabular form comparing and contrasting
- Relationship to geography

@@

>>>>

## Cartogram Visualisations of Disease

Cartograms utilise a map base, and distort using a statistic.

There was a long history that arrived cartogram maps.

@DMAHP
Thematic cartography began around 1800, spot maps, did not consider population rates
available data on the environment or society, weather crime rates. 
Disease mapping, evaluate geographical patterns, identify risk factors
For risks to be considered, need population for rates

@@

True maps, used for inspection to direct hypotheses.

@CIBMUK
-data exists within boundaries,
-areas within are shaded to match level of value,
-compare colour of areas to compare boundaries
-visual perception tends to weight the shades from the largest areas more heavily since they occupy more of our visual space
-display information in units of area, not population
-can mislead and bias inference

@@

### counts versus predictions
Need population, census data

@SEMA
 - 'Disease mapping is carried out to summarise spatial and spatio-temporal variation in risk.'
- Decriptive statistic of the relationship between geography and a disease.
- rapid visual summary of complex geographical information, descriptive purposes, to generate hypotheses, for surveillance, and to aid policy formation and resource allocation.
- population, exposure, and health data may have pointdata or aggregated summaries (countdata)
- potentially be subject to a number of inaccuracies
- diagnostic error or misclassification

- Not only the counts, but population, migration, deaths and births are modelled across census years.

@@

@DMAHP 
improve rates use age and sex
Standardised rates allow choropleth implementation

choice of variable:
relative risk (or equivalent), possibly in conjunction with statistical significance. Other functions included case frequencies or statistical significance alone. The choice of function makes an enormous difference to the appearance of the map, but the rationale for the choice adopted was rarely discussed.
The Finnish cancer atlas (Pukkala et al. 1987) was one of the earliest to use data smoothing, with a geometric centroid approach
More recently, a Canadian atlas of cancer incidence has appeared (Le et al. 1996; see Plate 1). Cancer is registered provincially, so considerable attention was paid to the quality and comparability of the data from contributing registries.
moving toward incidence

Very skeptical of smoothing

@HCUSMA The 1980s began computed generated cancer atlases

@@

@DMAHP
most disease atlases concerned cancer in developed countries
iso-demographic map: size by pop, colour by stat
@CIBMUK
'geographical differences does not automatically mean
that geography by itself makes the difference.'
Cartograms overcome this problem by standardizing the visual information presented to population size,

Issues: 'Many regions had such small expected frequencies that random variation would dominate the data. Interestingly, the smaller regions tended to occur in atlases that covered larger total populations.'
Smaller regions are a key issue, often occur when population has to be broken into many areas.

General introduction:
Cartograms stepped toward this goal by highlighting areas of interest. Made some areas bigger on the map space, kept relationship with neighbours.

UK: 
USA:
AUS: @DACQ AUS Cancer atlas example, reasons for choosing choropleth

@TVSSS Human cartography chapter
'in order to show the spatial structure of society the conventional use of maps of physical geography has to be rejected.'
strong opinions on lives of people not being accurately seen in geographical maps, not specifically disease related

only with the advent of sophisticated computer graphics,
visualization, has it become possible to do this on an easily replicable basis.
MAny examples and references to different map displays
### Contiguous

@CIBMUK 
-a trade-off between informational accuracy and intuitive representation

Importance of the statistic in this area being communicated was more important than keeping geography recognisable.

@@

## Comparison of

@CD 
discusses variable representation in maps
'Maps are designed following the cartographic visualization process'
Aknowledges map makers consider: methods and techniques, map type

Open source and commerical software can help make the maps, but decision still needs to be made, "nonprofessionals" can now make their own

(only worth including if it is possible for us to implement)
Tabular form comparing and contrasting
- Relationship to geography

@@

@DMAHP
"7. Increasing access to mapping technology The development of powerful computers has been a major force in the potential for disease mapping. It is likely that this trend will continue, so that larger databases and more rigorous analytical methods will become accessible to a wide circle of consumers. The availability of the European mortality atlas on the Internet is probably only the first such venture. In the future we will likely see capability for on-line manipulation of data to create maps to the user’s specifications, and an interface with other databases containing information on relevant covariables."

# References
